Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway

### 2022-01-24 analysis/paper/paper.Rmd
Final edits before returning to JAS:Reports

---
title: "Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: false
institute:
  - UiO: University of Oslo, Department of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as variable excavation practices, systems for artefact categorisation and raw-material availability, the sites chosen for analysis have all been excavated relatively recently and have a constrained geographical distribution. The assemblages were explored following two strains of analysis. The first of these entailed the use of artefact categories that are in established use within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely. The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods. Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Indicators (e.g. Clark and Barton 2017), originally devised for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature. This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting. Furthermore, this finding supports the notion that these measures can offer a powerful comparative tool in the analysis of lithic assemblages more generally.

keywords: |
  Mesolithic Scandinavia; Multivariate statistics; Mobility strategies; Whole Assemblage Behavioural Indicators
---
## Highlights {-}

@@

# Assemble plot of date ranges
temp_dist <- ggplot(datedata, aes(x = -mean_date, y = reorder(no, -as.numeric(no)), label = no)) +
  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = no, colour = excavation_project.x,
                   fill = excavation_project.x), size = 1.8) +
  geom_ridgeline(aes(x = date_grid, height = probabilities * 150,
                     fill = excavation_project.y, 
                     colour = excavation_project.x)) +
  scale_fill_manual(values = project_colors,
                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
  scale_colour_manual(values = project_colors,
                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
  labs(title = "B", fill = "Excavation project",
       x = "cal BCE", y = "Site number") +
  theme_classic() +
  theme(axis.ticks.y = element_blank(),
        legend.position = "bottom") +
  guides(fill =  guide_legend(title.position = "top", title.hjust = 0.5),
         colour = FALSE)

@@

The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_nonflint))`debitage and tool types. These represent artefact categories that have been used consistently across the reports. Consequently, sub-categories that have only been used in the classification of some inventories have been omitted. This for example pertains to what blanks have been used for the production of tools, and thus distinctions such as scrapers produced from flakes or blades, which has only been noted in some of the reports. Furthermore, the artefact data have been divided into flint and non-flint materials. Flint does not outcrop naturally in southern Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of climatic and geographical factors [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, for example by impacting accessibility by means of watercraft, the relatively constrained geographical distribution of the sites hopefully counteracts some environmentally given sources of variation.

@@

Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g. @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020]. Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [e.g. @breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g. @eigeland2015; @fuglestvedt2007; @mansrud2016]. These studies are, however, based on non-quantitative or uni- and bivariate methods, leaving the weighting of the many variables for the final interpretations unclear. To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.

The aim of the first of part of the analysis conducted here is to evaluate the degree to which the composition of the assemblages align with earlier studies that have employed more informal methods. This therefore assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021]. Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)). This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch which have been combined into the category 'small flint tools'. While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption of the largely intuitively determined aggregation procedure is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.

@@

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA. A version of the CA using the original artefact categories, as well as some additional configurations and ways to aggregate the variables are also available in the supplementary material to the paper.

@@

The general impression from the CA is that a chronological dimension accounts for a substantial amount of patterning in the data (Figure \@ref(fig:cor)). This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120; @lockyear2000]. The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)`% of the inertia or variance also means that the structure of the data is well-represented in the plots and that these therefore are likely reflect true patterning in the data.

The column plot reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. These assemblages are also to a larger extent characterised by core fragments, both in flint and non-flint materials, rather than cores. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both projectiles and non-flint blades are important constituents of the assemblages at these sites. The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages. Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that it is an early assemblage characterised by axe production in metarhyolite [@eigeland2014]. However, the site had been quite heavily impacted by modern disturbances that could have impacted the lithic material and which could explain its position as an outlier in the plot [@eigeland2014]. Finally, although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.

@@

```{r cor, fig.cap = "Correspondence analysis using the artefact count data. A) Row plot, B) Column plot. Points close together are more similar. By evaluating how the variables are distributed on the column plot it is possible to say how these define the two axes, in turn making it possible to relate the distribution of the sites in the row plot to the variables. As these are symmetrical plots, only general statements concerning the interrelation between the rows and the columns across the two plots can be made.", fig.height = 13}
# Plot sites
plt1 <- fviz_ca_row(calithics, col.row = "black",
                    fill = lithics$mean_date / 1000, title = "A",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

# Plot variables
plt2 <- fviz_ca_col(calithics, col.col = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "B") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")

# Plot sites for the flint data
plt3 <- fviz_ca_row(caflint, col.row = "black", fill = lithics$mean_date / 1000,
                    title = "A",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

# Variables for the flint data
plt4 <- fviz_ca_col(caflint, col.col = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "B") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")

@@

```{r flintcor, fig.height = 13, fig.cap = "Correspondence analysis using the flint data. A) Row plot, B) Column plot."}
plt3 /
plt4 
```

@@

The temporal trends associated with the curation index corresponds surprisingly well with trajectories of cultural development previously suggested in the literature, and does therefore, in my view, suggest that shifts in land-use and mobility patterns are the main drivers behind this empirical pattern---in line with the framework of @barton2011. Another perspective would be that this is not surprising at all [cf. @kuhn2015, 14], and that the previously demonstrated relevance of these measures across a wide range of contexts points to their pervasive relevance for the organisation of lithic technology, and, therefore, that there should be little reason to think Mesolithic south-eastern Norway should be any different. However, the conclusion that these these measures apply to and appear to capture the dimensions of interest in a relatively controlled empirical setting, reached by means of an exploratory analysis can only constitute a first analytical step. As @elster2015 [12] has pointed out, the human mind seems to have a propensity to settle for an explanation that *can* be true, as soon as this has been reached. This, however, can only constitute the absolute minimum of what is required of a proposed explanation. Subsequent steps should be to probe and challenge this explanatory framework, also in light of alternative hypotheses [e.g. @clark2009, 29--30; @perreault2019]. The empirical relationship does nonetheless hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond. Furthermore, the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with the Mesolithic of south-eastern Norway. The explicit quantification does, however, offer the possibility to conduct formal comparisons with a wide range of environmental, demographic and cultural dimensions across multiple scales of analysis.

@@

```{r supplementary1, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Row plot, B) Column plot.", fig.width = 10}
all_cats <- sites %>%
  # Make site names the rownames
  column_to_rownames(var = "no") %>% 
  # Deselect a few of the columns not be used
  select(-excavation_project, -excavated_m3, -dating_method)

@@

splt1 <- fviz_ca_row(caallcats, col.row = "black",
            fill = all_cats$mean_date / 1000, title = "A",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  ggtitle("Supplementary material") +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 23, face = "bold",
                                  family = "serif"))

splt2 <- fviz_ca_col(caallcats, col.col = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "B") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")

@@

```{r supplementary2, fig.height = 15, fig.cap = "Correspondence analysis collapsing artefact types irrespective of raw-material and including proportion of non-flint as its own variable. A) Row plot, B) Column plot."}

@@

splt3 <- fviz_ca_row(nonprop, col.row = "black",
            fill = all_nonflint_prop$mean_date / 1000, title = "A",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

splt4 <- fviz_ca_col(nonprop, col.col = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "B") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")

@@

```{r supplementary3, fig.height = 15, fig.cap = "Same as above, only that here the proportion of non-flint is used as a supplementary column. The negligble difference between this CA and that above indicates that the flint/non-flint distinction is integrated in the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated. These plots thus hide important variability that is captured in the ones presented in the main text. A) Row plot, B) Column plot."}

@@

splt5 <- fviz_ca_row(nonpropsup, col.row = "black",
            fill = all_nonflint_prop$mean_date / 1000, title = "A",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

# Seems to be a bug to change colour of supplementary column using 
# fviz_ca_row(): 
splt6 <- fviz(nonpropsup, element = "col", pointshape = 17,
              col.col.sup = "black",
             color = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "B") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")

>>>>

---
title: "Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: false
institute:
  - UiO: University of Oslo, Department of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as variable excavation practices, systems for artefact categorisation and raw-material availability, the sites chosen for analysis have all been excavated relatively recently and have a constrained geographical distribution. The assemblages were explored following two strains of analysis. The first of these entailed the use of artefact categories that are in established use within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely. The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods. Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Indicators (e.g. Clark and Barton 2017), originally devised for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature. This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting. Furthermore, this finding supports the notion that these measures can offer a powerful comparative tool in the analysis of lithic assemblages more generally.

keywords: |
  Mesolithic Scandinavia; Multivariate statistics; Mobility strategies; Whole Assemblage Behavioural Indicators
---

<!--- Page numbers in margin to avoid conflict with figure captions --->
\fancypagestyle{plain}{%
  \fancyhf{}
  \fancyfoot[R]{\leavevmode\rlap{\thepage~}}
}
\pagestyle{plain}

## Highlights {-}

@@

# Assemble plot of date ranges
temp_dist <- ggplot(datedata, aes(x = -mean_date, y = reorder(no, -as.numeric(no)), label = no)) +
  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = no, colour = excavation_project.x,
                   fill = excavation_project.x), size = 1.8) +
  geom_ridgeline(aes(x = date_grid, height = probabilities * 150,
                     fill = excavation_project.y, 
                     colour = excavation_project.x)) +
  scale_fill_manual(values = project_colors,
                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
  scale_colour_manual(values = project_colors,
                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
  labs(title = "B", fill = "Excavation project",
       x = "cal BCE", y = "Site number") +
  theme_classic() +
  theme(axis.ticks.y = element_blank(),
        legend.position = "bottom") +
  guides(fill =  guide_legend(title.position = "top", title.hjust = 0.5),
         colour = "none")

@@

The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_nonflint))` debitage and tool types. These represent artefact categories that have been used consistently across the reports. Consequently, sub-categories that have only been used in the classification of some inventories have been omitted. This for example pertains to what blanks have been used for the production of formal tools, which has only been noted in some of the reports. Furthermore, the artefact data have been divided into flint and non-flint materials. Flint does not outcrop naturally in southern Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of climatic and geographical factors [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, for example by impacting accessibility by means of watercraft, the relatively constrained geographical distribution of the sites hopefully counteracts some environmentally given sources of variation.

@@

Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g. @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020]. Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [e.g.  @breivik2020; @breivik2016; @viken2018c; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g. @eigeland2015; @fuglestvedt2007; @mansrud2016]. These studies are, however, based on non-quantitative or descriptive uni- and bivariate methods, leaving the weighting of the many variables for the final interpretations unclear. To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.

The aim of the first of part of the analysis conducted here is to evaluate the degree to which the composition of the assemblages align with earlier studies that have employed more informal methods. This therefore assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021]. Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)). Artefact categories believed not to be internally consistent and categorically exclusive have therefore been collapsed for the analysis, as their contribution as discrete analytical units could potentially be misleading. An underlying assumption of the largely intuitively determined aggregation procedure is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While aggregating artefact categories in this manner could subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.

@@

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports [e.g. @baxter1994; @shennan1997]. As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA. A version of the CA using the original artefact categories, as well as some additional configurations and ways to aggregate the variables are also available in the supplementary material to the paper.

@@

The general impression from the CA is that a chronological dimension accounts for a substantial amount of patterning in the data (Figure \@ref(fig:cor)). This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)B), as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)A, @baxter1994, 119--120; @lockyear2000]. The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)`% of the inertia or variance also means that the structure of the data is well-represented in the plots and that these therefore are likely reflect true patterning in the data.

The column plot reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. These assemblages are also to a larger extent characterised by core fragments, both in flint and non-flint materials, rather than cores. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both projectiles and non-flint blades are important constituents of the assemblages at these sites. The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages. Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that it is an early assemblage characterised by axe production in metarhyolite [@eigeland2014]. However, the site had been quite heavily impacted by modern disturbances that could have impacted the lithic material and which could explain its position as an outlier in the plot. Finally, although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.

@@

```{r cor, fig.cap = "Correspondence analysis using the artefact count data. A) Column plot (variables), B) Row plot (sites). Points close together are more similar. By evaluating how the variables are distributed on the column plot it is possible to say how these define the two axes, in turn making it possible to relate the distribution of the sites in the row plot to the variables. As these are symmetrical plots, only general statements concerning the interrelation between the rows and the columns across the two plots can be made.", fig.height = 13}
# Plot variables
plt1 <- fviz_ca_col(calithics, col.col = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "A") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")

# Plot sites
plt2 <- fviz_ca_row(calithics, col.row = "black",
                    fill = lithics$mean_date / 1000, title = "B",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

# Variables for the flint data
plt3 <- fviz_ca_col(caflint, col.col = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "A") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")

# Sites for the flint data
plt4 <- fviz_ca_row(caflint, col.row = "black", fill = lithics$mean_date / 1000,
                    title = "B",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

@@

```{r flintcor, fig.height = 13, fig.cap = "Correspondence analysis using the flint data. A) Column plot, B) Row plot."}
plt3 /
plt4 
```

@@

The temporal trends associated with the curation index corresponds surprisingly well with trajectories of cultural development previously suggested in the literature, and does therefore, in my view, suggest that shifts in land-use and mobility patterns are the main drivers behind this empirical pattern---in line with the framework of @barton2011. Another perspective would be that this is not surprising at all [cf. @kuhn2015, 14], and that the previously demonstrated relevance of these measures across a wide range of contexts points to their pervasive relevance for the organisation of lithic technology, and, therefore, that there should be little reason to think Mesolithic south-eastern Norway should be any different. However, the conclusion that these these measures apply to and appear to capture the dimensions of interest in a relatively controlled empirical setting, reached by means of an exploratory analysis can only constitute a first analytical step. As @elster2015 [12] has pointed out, the human mind seems to have a propensity to settle for an explanation that *can* be true, as soon as this has been reached. This, however, can only constitute the absolute minimum of what is required of a proposed explanation. Subsequent steps should be to probe and challenge this explanatory framework, also in light of alternative hypotheses [e.g. @clark2009, 29--30; @perreault2019, 1--22]. The empirical relationship does nonetheless hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond. Furthermore, the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with the Mesolithic of south-eastern Norway. The explicit quantification does, however, offer the possibility to conduct formal comparisons with a wide range of environmental, demographic and cultural dimensions across multiple scales of analysis.

@@

```{r supplementary1, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Column plot, B) Row plot.", fig.width = 10}
all_cats <- sites %>%
  # Make site names the rownames
  column_to_rownames(var = "no") %>% 
  # Deselect a few of the columns not be used
  select(-excavation_project, -excavated_m3, -dating_method)

@@

splt1 <- fviz_ca_col(caallcats, col.col = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE) +
  labs(colour = "Contribution to total inertia") +
  ggtitle("Supplementary material") +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 23, face = "bold",
                                  family = "serif"))

splt2 <- fviz_ca_row(caallcats, col.row = "black",
            fill = all_cats$mean_date / 1000, title = "B",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

@@

```{r supplementary2, fig.height = 15, fig.cap = "Correspondence analysis collapsing artefact types irrespective of raw-material and including proportion of non-flint as its own variable.  A) Column plot, B) Row plot."}

@@

splt3 <- fviz_ca_col(nonprop, col.col = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "A") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")

splt4 <- fviz_ca_row(nonprop, col.row = "black",
            fill = all_nonflint_prop$mean_date / 1000, title = "B",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

@@

```{r supplementary3, fig.height = 15, fig.cap = "Same as above, only that here the proportion of non-flint is used as a supplementary column. A) Column plot, B) Row plot. The negligble difference between this CA and that above indicates that the flint/non-flint distinction is integrated in the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated. These plots thus hide important variability that is captured in the ones presented in the main text."}

@@

# Seems to be a bug to change colour of supplementary column using 
# fviz_ca_row(): 
splt5 <- fviz(nonpropsup, element = "col", pointshape = 17,
              col.col.sup = "black",
             color = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "A") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")

splt6 <- fviz_ca_row(nonpropsup, col.row = "black",
            fill = all_nonflint_prop$mean_date / 1000, title = "B",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")
